---
title: "ODBA_model"
output: html_document
date: "2024-05-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, message= FALSE, warning=FALSE}
library(dplyr)
library(corrplot)
library(lme4) # for lmer
library(nlme) #for gls (glmm)
library(lmerTest)
library(Matrix) 
library(MuMIn) # for Dredge()
library(car) # for vif()
library(lubridate)
library(GGally) #correlation synthese
library(mgcv) #for gam
library(lmtest) #Breusch-Pagan test
library(effects)
library(flexmix) #for BIC extraction
library(performance) #R²
library(gridExtra)#drif.arrange
```

## Prepare DATA

```{r}
# data1 <- read.table("D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/breeder_all_ok.txt", header = T, dec = ",")
# data <- data1[,c("device", "id", "Deviceyear",  "POP", "SEX", "daily_mean_ODBA", "PRODUCTION_envol", "CHICKAGE", "CUMUL_prcp_ann_pond", "SPEI_har_pond", "SPI_pond", "ecart_cumul_annuel_moyen")]

data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025_odba.txt",
    header = T,
    dec = ","
  )
#79 periods
colnames(data1)[colnames(data1) == "SEXE"] <- "SEX"
data1$daily_mean_ODBA <- data1$ODBA.mG.#ligne à retirer si voulu en G


#Convert into the good type
data1 <- data1 %>%
  mutate(across(c(Deviceyear, id_30, POP, SEX, PRODUCTION_envol, CHICKAGE, device), as.factor))

data1$SPEI_har_pond <- as.numeric(data1$SPEI_har_pond)
#no need to scale
data <- data1

table(data$PRODUCTION_envol)
# Only one individual has 5 chicks :add it to the 4+ class
data$PRODUCTION_envol <- ifelse(data$PRODUCTION_envol == 4 | data$PRODUCTION_envol == 5, "4+", as.character(data$PRODUCTION_envol))
data <- data %>%
  mutate(across(c(POP, SEX, PRODUCTION_envol, CHICKAGE), as.factor)) %>%
  mutate(across(c(SPEI_har_pond, daily_mean_ODBA), as.numeric))

summary(data)
```

# Y distribution

```{r}
test <- data
# test$daily_mean_ODBA1 <- test$daily_mean_ODBA
# test$daily_mean_ODBA <- test$daily_mean_ODBA/1000 #en G au lieu de mG
summary(test)
test$daily_mean_ODBA_log <- log(test$daily_mean_ODBA)#si besoin

#Check for missing values
colSums(is.na(test))

#Y distribution

par(mfrow=c(2,2))
# Boxplot
boxplot(test$daily_mean_ODBA,col='blue',ylab='Monthly mean of the daily mean ODBA')
# Cleveland plot
dotchart(test$daily_mean_ODBA,pch=16,col='blue',xlab='Monthly mean of the daily mean ODBA')
# Histogram
hist(test$daily_mean_ODBA,col='blue',xlab="Monthly mean of the daily mean ODBA",main="")
# Quantile-Quantile plot
qqnorm(test$daily_mean_ODBA,pch=16,col='blue',xlab='')
qqline(test$daily_mean_ODBA,col='red')
# Non normal
shapiro.test(test$daily_mean_ODBA) #W = 0.96351, p-value = 0.02343 Non normal law
#No outlier
```

### Potential relationships between Xs and Y

```{r}
par(mfrow=c(1,4))

#PRCP
# SPEI
plot(test$daily_mean_ODBA~test$SPEI_har_pond,pch=16,col='blue',xlab='SPEI',ylab='Monthly mean of the daily Odba')

# Production
boxplot(daily_mean_ODBA ~ PRODUCTION_envol, data = test, main = "Monthly mean of the daily Odba")
# sex
boxplot(daily_mean_ODBA ~ SEX, data = test, col = c('blue3','red2'), main = "Monthly mean of the daily Odba")
# Chick age
boxplot(daily_mean_ODBA ~ CHICKAGE, data = test, col = c('blue3','red2'), main = "Monthly mean of the daily Odba")
```

Mean daily ODBA: increases with the number of chicks, decreases in males, increases with the chick age

```{r}
### Choice of the random factors
#LM, constant model
mod <- lm(daily_mean_ODBA ~1, data = test)
AICc(mod) #-224.0653

#LMM
mod1 <- lmer(daily_mean_ODBA ~(1|POP/device), data = test)
AICc(mod1)#-213.4173
mod2 <- lmer(daily_mean_ODBA ~(1|POP/device) + (1|PRODUCTION_envol), data = test)
AICc(mod2)#-213.4173
mod3 <- lmer(daily_mean_ODBA ~(1|PRODUCTION_envol), data = test)
AICc(mod3) #-214.6008 
mod4 <- lmer(daily_mean_ODBA ~(1|device), data = test)
AICc(mod4) #-215.2044*
mod5 <- lmer(daily_mean_ODBA ~(1|device) + (1|POP), data = test)
AICc(mod5)#-213.4173
mod6 <- lmer(daily_mean_ODBA ~(1|POP), data = test)
AICc(mod6)#-214.9878
mod7 <- lmer(daily_mean_ODBA ~(1|device) + (1|PRODUCTION_envol), data = test)
AICc(mod7) #-213.7291

#Compare AICc
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6)
aiccs <- sapply(model_list, AICc)
aiccs #-224.0653 -213.4173 -211.9533 -214.6008 -215.2044 -213.4173 -214.9878
best_model <- model_list[[which.min(aiccs)]] 
best_model # daily_mean_ODBA ~ (1 | device)
#not a lot of diference  with pop but the individual variability incorporates the population variability

```

Best model with only device as random factor

### GAMM with the complete model: linearity of the relationship?

```{r}
modgam <- gamm(daily_mean_ODBA ~ s(SPEI_har_pond)+ SEX + CHICKAGE,  random = list(device =~1), data = test)
summary(modgam$gam)
AICc(modgam)# -267.1109

#remove SEX
modgam4 <- gamm(daily_mean_ODBA ~ s(SPEI_har_pond) + CHICKAGE,  random = list(device =~1), data = test)
AICc(modgam4)#-262.4902 No, the model is worst without it

#remove Chick age
modgam4 <- gamm(daily_mean_ODBA ~ s(SPEI_har_pond) + SEX,  random = list(device =~1), data = test)
AICc(modgam4) # -225.7062, keep both factors

#best model
modgam <- gamm(daily_mean_ODBA ~ s(SPEI_har_pond) + SEX + CHICKAGE,  random = list(device =~1), data = test)
AICc(modgam)

plot(modgam$gam)
plot(modgam$lme)
summary(modgam$gam) #Ref edf = 1 linear relationship

 # Each plot represents the smoothed effect of a continuous variable on the response variable. 
 # The solid line is the estimated effect, and the shaded areas represent the 95% confidence intervals.
summary(modgam$lme)
```

Linear relationships --\> LMM or GLMM

### LMM Model selection

```{r,results='hide',fig.keep='all'}

#Complete model
modglm <- lmer(daily_mean_ODBA~ SPEI_har_pond + SEX + CHICKAGE +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)# -269.5217 
r.squaredGLMM(modglm)


##FACTORS
#remove sex
modglm <- lmer(daily_mean_ODBA~ SPEI_har_pond + CHICKAGE + (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #-264.8349, the model is worse, keep it
r.squaredGLMM(modglm)

#remove chick age
modglm <- lmer(daily_mean_ODBA~ SPEI_har_pond + SEX + (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) # -226.9194 idem keep them both
r.squaredGLMM(modglm)

#SPEI alone
modglm <- lmer(daily_mean_ODBA~ SPEI_har_pond + (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #-222.4329
r.squaredGLMM(modglm)

#only chickage
modglm <- lmer(daily_mean_ODBA~ CHICKAGE + (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #827.3163
r.squaredGLMM(modglm)

#only sex
modglm <- lmer(daily_mean_ODBA~ SEX + (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #-222.4329
r.squaredGLMM(modglm)

#chickage et sex
modglm <- lmer(daily_mean_ODBA~ SEX + CHICKAGE + (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #-222.4329
r.squaredGLMM(modglm)

#NULL
modglm <- lmer(daily_mean_ODBA~ (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #-223.1895
r.squaredGLMM(modglm)

modok <- lmer(daily_mean_ODBA~ SEX + CHICKAGE + (1|device), data = test, REML = T, na.action = na.fail)

```
Avec Dredge

```{r}
fullmod <-  lmer(daily_mean_ODBA~ SPEI_har_pond + SEX + CHICKAGE +
                  (1|device), data = test, REML = F, na.action = na.fail)

Dredge <- dredge(fullmod)

#2 modèles en concurrence
modglm <- lmer(daily_mean_ODBA~ SEX + CHICKAGE +
                  (1|device), data = test, REML = T, na.action = na.fail)
#Estimate nul pour ajout de SPEI + CHICKAGE + SEX ?
modglm <- lmer(daily_mean_ODBA~ SPEI_har_pond + SEX + CHICKAGE +
                  (1|device), data = test, REML = T, na.action = na.fail)
confint(modglm, oldNames=FALSE) #CHICKAGE ESTIMATE NULL
summary(modglm)

```

### Model validation

```{r}
#selected model
#REML = T for model validation (provides less biased estimates of the variance components)
modok <- lmer(daily_mean_ODBA~ SEX + CHICKAGE + (1|device), data = test, REML = T, na.action = na.fail)

#residuals (normality and independance)
qqnorm(resid(modok))
qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test$CHICKAGE,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Chick age",
         main = "")
abline(h = 0)

plot(residuals(modok)~ test$SEX,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SEX",
         main = "")
abline(h = 0)

plot(residuals(modok)~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

plot(Effect("CHICKAGE",modok,partial.residuals=TRUE))
plot(Effect("SEX",modok,partial.residuals=TRUE))

#Multicolinearity
vif(modok)
#(Zuur et al: threshold 5-6)

#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#        R2m       R2c
#[1,] 0.4640072 0.5542994

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

BIC(modok)

summary(modok)

#                Estimate Std. Error        df t value Pr(>|t|)    
# (Intercept)    0.275286   0.010271 21.560931  26.802  < 2e-16 ***
# SPEI_har_pond -0.008689   0.006121 71.090867  -1.420   0.1601    
# SEXM          -0.034594   0.012218 16.715687  -2.831   0.0117 *  
# CHICKAGE1      0.069207   0.008716 59.599359   7.940 6.47e-11 ***

###PLOT
library(ggeffects)
effect_data <- ggpredict(modok, terms = c("SPEI_har_pond", "SEX", "CHICKAGE"), ci_level = 0.95)

ggplot(effect_data, aes(x = x, y = predicted, group = group)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.15) +
  geom_line(aes(color = group), alpha = 2) +
  scale_color_manual(values = c("darkred", "deepskyblue4")) +  # Les lignes noires pour les deux groupes
  scale_fill_manual(values = c("M" = "deepskyblue", "F" = "red")) +
  facet_wrap(~facet, scales = "free") +
  labs(title = "Effect of SPEI, SEX, and CHICKAGE on Daily Mean ODBA",
       x = "SPEI_har_pond", y = "Daily Mean ODBA") +
  theme_classic()



#FA méthod
coef <- coefficients(modok)
coef
coef(modok)


library(emmeans)
## Obtenir les moyennes marginales estimées
emmeans_result <- emmeans(modok, ~ SEX + CHICKAGE)
#Compute estimated marginal means (EMMs) for specified factors or factor combinations in a linear model; and optionally, comparisons or contrasts among them. EMMs are also known as least-squares means.

# Convertir en data.frame pour ggplot
emmeans_df <- as.data.frame(emmeans_result)


ggplot(test, aes(x = CHICKAGE, y = daily_mean_ODBA, fill = SEX)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.6) +
  labs(x = "Sex", y = "Daily mean ODBA",
       fill = "Period") +
  theme_minimal()

ggplot(test, aes(x = as.factor(CHICKAGE), y = daily_mean_ODBA, fill = SEX)) +
  geom_boxplot(position = position_dodge(width = 0.8), 
               alpha = 0.3,  # Augmente la transparence
               color = "black") +  # Contour noir
  geom_jitter(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.8), 
              size = 0.8, color = "black") +  # Points noirs
  labs(x = "Chick Age", y = "Daily Mean ODBA",
       fill = "Sex") +  # Ajustement des légendes
  theme_classic()

ggplot(test, aes(x = CHICKAGE, y = daily_mean_ODBA, fill = SEX)) +
  geom_errorbar(data = emmeans_df, aes(x = as.factor(CHICKAGE), y = emmean, 
                                       ymin = lower.CL, ymax = upper.CL, color = SEX),
                position = position_dodge(width = 0.8), width = 0.4) +  # Barres d'erreur
  geom_point(data = emmeans_df, aes(x = as.factor(CHICKAGE), y = emmean, color = SEX), 
             position = position_dodge(width = 0.8), size = 3) +  # Point pour la moyenne
  labs(x = "Sex", y = "Daily mean ODBA",
       fill = "Period") +
  theme_minimal()


# Ajouter les moyennes ajustées issues de emmeans
ggplot(test, aes(x = SEX, y = daily_mean_ODBA, fill = CHICKAGE)) +
  geom_boxplot(position = position_dodge(width = 0.8), alpha = 0.5) +
  geom_point(data = emmeans_df, aes(x = SEX, y = emmean, color = CHICKAGE),
             position = position_dodge(width = 0.8), size = 3, shape = 21) +
  geom_errorbar(data = emmeans_df, aes(x = SEX, y = emmean, ymin = emmean - SE, ymax = emmean + SE, color = CHICKAGE),
                position = position_dodge(width = 0.8), width = 0.2) +
  labs(x = "Sex", y = "Daily mean ODBA",
       fill = "Period", color = "Period") +
  theme_minimal()

#attention ici SE pas IC95 je crois ?!

```

# Représentation 2

```{r}
modok <- lmer(daily_mean_ODBA~   SPEI_har_pond + SEX + CHICKAGE + (1|device), data = test, REML = T, na.action = na.fail)

library(nlme)
modok<-lme(daily_mean_ODBA~ SPEI_har_pond + SEX + CHICKAGE, 
                 random = ~1|device, data = test, na.action = na.fail, method = "REML")



newdat <- expand.grid(SPEI_har_pond=seq(-1.569,1.86,0.1), SEX = c("F", "M"), CHICKAGE = c("0","1"))
library(stats)
newdat$pred <- stats::predict(modok, newdat, level = 0)

Designmat <- model.matrix(formula(modok)[-2], newdat)
predvar <- diag(Designmat %*% vcov(modok) %*% t(Designmat)) 
newdat$SE <- sqrt(predvar) 
newdat$SE2 <- sqrt(predvar+modok$sigma^2)

cmult <- 1.96

g0 <- ggplot(newdat, aes(x = SPEI_har_pond, y = pred, colour = SEX)) + 
  geom_line() +
  geom_ribbon(aes(ymin = pred - cmult * SE, ymax = pred + cmult * SE, fill = SEX), alpha = 0.1) +
  geom_point(data = test, aes(x = SPEI_har_pond, y = daily_mean_ODBA, col = SEX)) +
  scale_color_manual(values = c("darkred", "deepskyblue4")) +
  scale_fill_manual(values = c("M" = "deepskyblue", "F" = "red")) +
  facet_wrap(~ CHICKAGE, scales = "fixed", ncol = 1) +
  theme_classic() +
  labs(x = "SPEI", y = "ODBA") +
  theme(
    axis.text.x = element_text(size = 16),  # Increase size of x-axis text
    axis.text.y = element_text(size = 16),  # Increase size of y-axis text
    axis.title.x = element_text(size = 18), # Increase size of x-axis title
    axis.title.y = element_text(size = 18), # Increase size of y-axis title
    strip.text = element_blank()  # Remove facet labels
  )

```


### Model validation: remove the 4 incomplete months

```{r}
# #remove the 4 incomplete months
# test1 <- test %>%
#   filter(!(id %in% c("6_21_2", "17_23_2", "25_20_1", "37_20_1")))
# 
# 
# #REML = T for model validation (provides less biased estimates of the variance components)
# modok <- lmer(daily_mean_ODBA~  ecart_cumul_annuel_moyen + SEX + CHICKAGE + (1|device), data = test1, REML = T, na.action = na.fail)
# 
# 
# #residuals (normality and independance)
# qqnorm(resid(modok))
# qqline(resid(modok)) #ok for residuals normality
# #idem
# hist(residuals(modok))
# 
# # residuals vs fitted (homoscedasticity)
# plot(residuals(modok)~fitted(modok)
#       , col='blue'
#       , pch=16)
# abline(h = 0)
# #ok for independance: no pattern
# 
# 
# #residuals against factors (relationship linearity and variance equality)
# plot(residuals(modok)~ test1$CHICKAGE,
#          varwidth = TRUE,
#          ylab = "Residuals",
#          xlab = "Chick age",
#          main = "")
# abline(h = 0)
# 
# plot(residuals(modok)~ test1$SEX,
#          varwidth = TRUE,
#          ylab = "Residuals",
#          xlab = "SEX",
#          main = "")
# abline(h = 0)
# 
# plot(residuals(modok)~ test1$ecart_cumul_annuel_moyen,
#          ylab = "Residuals",
#          xlab = "Anomaly in mean annual prcp",
#          main = "")
# abline(h = 0)
# 
# plot(Effect("ecart_cumul_annuel_moyen",modok,partial.residuals=TRUE))
# 
# #Multicolinearity
# vif(modok)
# #(Zuur et al: threshold 5-6)
# # ecart_cumul_annuel_moyen                      SEX                 CHICKAGE 
# #                 1.007907                 1.003276                 1.008072
# 
# #marginal and conditional R-squared values 
# r.squaredGLMM(modok)
# #  0.4538513 0.5361808
# 
# #IConfidence interval (95%) for each variable
# confint(modok, oldNames=FALSE)
# 
# 
# #Homoscedasticity Breusch-Pagan test
# # Model residuals
# residus <- resid(modok)
# # bptest(resid)
# 
# #Regression residuals - variables
# lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test1)
# 
# # Effectuer le test Breusch-Pagan
# bptest(lm_resid)
# #ok !
# 
# BIC(modok)
# 
# summary(modok) #same effects
# 
# #                  Estimate Std. Error      df t value Pr(>|t|)    
# # (Intercept)               275.159      9.863  25.087  27.897  < 2e-16 ***
# # ecart_cumul_annuel_moyen   -9.691      4.765  71.565  -2.034   0.0457 *  
# # SEXM                      -33.729     11.774  17.926  -2.865   0.0103 *  
# # CHICKAGE2                  63.396      8.851  57.838   7.163 1.59e-09 ***
```

Conclusion: The monthly mean of the daily ODBA: - decreases when prcp departure to the mean is high so in wetter conditions - decreases in males - increases when the chick is older Hypotheses: drought + old chick = more foraging and flying activities

Test du modèle en enlevant les mois de - de 21j

```{r}
id_30_to_delete <- c("190657_2021_2", "190665_2020_1", "200015_2020_1")
test1 <-  test[!(test$id_30 %in% id_30_to_delete), ]


#selected model
#REML = T for model validation (provides less biased estimates of the variance components)
modok <- lmer(daily_mean_ODBA~   SPEI_har_pond + SEX + CHICKAGE + (1|device), data = test1, REML = T, na.action = na.fail)
AICc(modok)#798.3452 ->770.3278

#residuals (normality and independance)
qqnorm(resid(modok))

qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test1$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#            R2m       R2c
# [1,] 0.4549217 0.5315197

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test1)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok ! p>0.005

summary(modok)

#PLot it
plot(residuals(modok)~ test1$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

```

