---
title: '%foraging'
output: html_document
date: "2024-05-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, message= FALSE, warning=FALSE}
library(dplyr)
library(corrplot)
library(lme4) # for lmer
library(nlme) #for gls (glmm)
library(lmerTest)
library(Matrix) 
library(MuMIn) # for Dredge()
library(car) # for vif()
library(lubridate)
library(GGally) #correlation synthese
library(mgcv) #for gam
library(lmtest) #Breusch-Pagan test
library(effects)
library(flexmix) #for BIC extraction
library(performance) #R²
library(gridExtra)#drif.arrange
```

## Prepare DATA

```{r}

data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025.txt",
    header = T,
    dec = ","
  )
#79 periods

colnames(data1)[colnames(data1) == "SEXE"] <- "SEX"
colnames(data1)[colnames(data1) == "n_all_bhv"] <- "tot_month_activity"
colnames(data1)[colnames(data1) == "n_foraging"] <- "tot_month_foraging"
colnames(data1)[colnames(data1) == "prop_for_month"] <- "mean_.foraging"


data <-
  data1[, c(
    "device",
    "ID",
    "id_30",
    "mean_.foraging",
    "Deviceyear",
    "tot_month_activity",
    "tot_month_foraging",
    "POP",
    "SEX",
    "PRODUCTION_envol",
    "CHICKAGE",
    "SPEI_har_pond"
  )]


#Convert into the good type
data <- data %>%
  mutate(across(c(POP, SEX, PRODUCTION_envol, CHICKAGE, id_30, device), as.factor))
  #standardize the numeric environmental variables
data$SPEI_har_pond <- as.numeric(data$SPEI_har_pond)
data$mean_.foraging <- as.numeric(data$mean_.foraging)

#no scale

#rename columns
data <- data %>%
  rename(prop_foraging = mean_.foraging)
table(data$PRODUCTION_envol)
# Only one individual has 5 chicks :add it to the 4+ class
data$PRODUCTION_envol <-
  ifelse(
    data$PRODUCTION_envol == 4 |
      data$PRODUCTION_envol == 5,
    "4+",
    as.character(data$PRODUCTION_envol)
  )
data$PRODUCTION_envol <- as.factor(data$PRODUCTION_envol)
summary(data)
```

# % Foraging

#How to explain the monthly mean of the daily % of foraging activity?

### Prepare data

```{r}
test <-data
 
#Check for missing values
colSums(is.na(test))

#Y distribution
#Y between 0 and 1 (%)
test$prop_foraging <- test$prop_foraging/100
ratio <- cbind(test$tot_month_foraging, test$tot_month_activity-test$tot_month_foraging) #for the glm building

par(mfrow=c(2,2))
# Boxplot
boxplot(test$prop_foraging,col='blue',ylab='Daily %of foraging')
# Cleveland plot
dotchart(test$prop_foraging,pch=16,col='blue',xlab='Monthly mean \n of the daily %of foraging')
# Histogram
hist(test$prop_foraging,col='blue',xlab="Monthly mean\n of the daily %of foraging",main="")
# Quantile-Quantile plot
qqnorm(test$prop_foraging,pch=16,col='blue',xlab='')
qqline(test$prop_foraging,col='red')
# Non normal
shapiro.test(test$prop_foraging) #W = 0.97681, p-value = 0.1598 Normal law ok
#No outlier
```

### Potential relationships between Xs and Y

```{r}
par(mfrow=c(1,4))

# SPEI
plot(
  test$prop_foraging ~ test$SPEI_har_pond,
  pch = 16,
  col = 'blue',
  xlab = 'SPEI',
  ylab = 'Monthly mean of the daily %of foraging'
)

# Production
boxplot(prop_foraging ~ PRODUCTION_envol,
        data = test,
        main = "Monthly mean of the daily %of foraging")
# sex
boxplot(
  prop_foraging ~ SEX,
  data = test,
  col = c('blue3', 'red2'),
  main = "Monthly mean of the daily %of foraging"
)
# Chick age
boxplot(
  prop_foraging ~ CHICKAGE,
  data = test,
  col = c('blue3', 'red2'),
  main = "Monthly mean of the daily %of foragingt"
)
```

Mean daily % foraging: increases with the number of chicks, decreases in males, increases with the chick age

### Choice of the random factors

```{r}
#GLM, constant model
mod <- glm(ratio~1, family = binomial(link = "logit"), data = test)
AICc(mod)#3883.344
#GLMM avec ratio
mod0 <- glmer(ratio~(1|POP/device),family = binomial(link = "logit"), data = test)
AICc(mod0)#3056.669
#singular
mod1 <- glmer(ratio~(1|POP),family = binomial(link = "logit"), data = test)
AICc(mod1)#3885.45
mod2 <- glmer(ratio~(1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test)
AICc(mod2)#2948.93
mod3 <- glmer(ratio~(1|PRODUCTION_envol),family = binomial(link = "logit"), data = test)
AICc(mod3)#3633.468
mod4 <- glmer(ratio~(1|device),family = binomial(link = "logit"), data = test)
AICc(mod4)#3054.507
mod5 <- glmer(ratio~(1|device) + (1|POP),family = binomial(link = "logit"), data = test)
AICc(mod5)#3056.669

#Compare AICc
model_list <- list(mod,mod0, mod1, mod2, mod3, mod4, mod5)
aiccs <- sapply(model_list, AICc)
aiccs #3883.344 3056.669 3885.450 2948.930 3633.468 3054.507 3056.669
best_model <- model_list[[ which.min(aiccs)]]
best_model #ratio ~ (1 | POP/device) + (1 | PRODUCTION_envol)
```

Keep ratio \~ (1 \| POP/device) + (1 \| PRODUCTION_envol)

### GAMM with the complete model: linearity of the relationship?

```{r}
#####Continuous variables 

#remove ecart
modgam2 <- gamm(ratio ~ s(SPEI_har_pond) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modgam2) #657.2088

#########Factors
#remove sex
modgam4 <- gamm(ratio ~   s(SPEI_har_pond) + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modgam4) #935.2852 keept it

#remove chick age
modgam5 <- gamm(ratio ~   s(SPEI_har_pond) + SEX,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modgam5) #1268.581
#Keep the factors

modgam <-  gamm(ratio ~ s(SPEI_har_pond) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modgam)#657.2088

par(mfrow =c(2,2))
plot(modgam$gam)
summary(modgam$gam) #REf df = 8.628 - complex relationship

 # Each plot represents the smoothed effect of a continuous variable on the response variable. 
 # The solid line is the estimated effect, and the shaded areas represent the 95% confidence intervals.
summary(modgam$lme)
```
## GAMM model:

```{r}
#Selected model
modgam <- gamm(ratio ~ s(SPEI_har_pond) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modgam)#657.2088

#Standardized residuals vs predicted values: random distribution = ok
residuals_gam <- residuals(modgam$gam, type = "pearson")
fitted_values_gam <- fitted(modgam$gam)

plot(fitted_values_gam, residuals_gam, xlab = "Predicted values", ylab = "Standardized residuals",
     main = "Standardized residuals vs predicted values")
abline(h = 0, col = "red") #ok

plot(modgam$gam)
```

### Model validation

```{r}
modok <- gamm(ratio ~ s(SPEI_har_pond) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok)#657.2088

#residuals
residu <- residuals(modok$gam, type = "pearson")
fitted <- fitted(modok$gam)
qqnorm(residu)
qqline(residu)
#normalité des résidus
hist(residu)

modok_gam <- modok$gam
# vif(modok_gam)
#(Zuur et al: non-linear relationships are not picked up by the VIF

# residuals vs fitted (homoscedasticity)
plot(residu~fitted
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residu~ test$CHICKAGE,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Chick age",
         main = "")
abline(h = 0)

plot(residu~ test$SEX,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SEX",
         main = "")
abline(h = 0)

plot(residu~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)


#R²
summary(modok$gam)#R-sq.(adj) =  0.606

#BIC
BIC(modok$lme)# 672.2175
#AICc
AICc(modok)#657.2088

#is k sufficient?
gam.check(modok_gam)
#p-value >0.05 = yes

#IConfidence interval (95%) for each variable
confint(modok_gam, oldNames=FALSE)
plot(modok$gam) #hard to interpret, is k really ok and not too high?
summary(modok$gam)
```

#K selection

```{r}
#Routine to visually compare the different values of k

# Function to fit the model and extract the smooth term plot data
fit_gamm_and_extract_plot_data <- function(k1, data) {
  mod <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k = k1) + 
                        SEX + CHICKAGE, 
              random = list(POP = ~1|device, PRODUCTION_envol = ~1), 
              family = binomial(link = "logit"), data = data)
  
  # Extract smooth terms
  smooth_terms <- predict(mod$gam, type = "terms", se.fit = TRUE)
  df <- data.frame(data, 
                   SPEI_har_pond_smooth = smooth_terms$fit[, "s(SPEI_har_pond)"], 
                   SPEI_har_pond_se = smooth_terms$se.fit[, "s(SPEI_har_pond)"], 
                   k1 = k1)
  
  # Extract AICc value
  aicc_value <- AICc(mod$lme)
  
  list(plot_data = df, aicc = aicc_value)
}

# Fit models for k values from 1 to 9 and store results
k_values <- 1:6
results <- lapply(k_values, function(k) fit_gamm_and_extract_plot_data(k, test))

# Combine plot data
plot_data <- do.call(rbind, lapply(results, function(res) res$plot_data))

# Extract AICc values
aicc_values <- data.frame(k = k_values, AICc = sapply(results, function(res) res$aicc))

# Print AICc values for comparison
print(aicc_values)

# Plot SPEI_har_pond smooth terms
p1 <- ggplot(plot_data, aes(x = SPEI_har_pond, y = SPEI_har_pond_smooth, color = as.factor(k1))) +
  geom_line() +
  geom_line(aes(y = SPEI_har_pond_smooth + 2 * SPEI_har_pond_se), linetype = "dashed") +
  geom_line(aes(y = SPEI_har_pond_smooth - 2 * SPEI_har_pond_se), linetype = "dashed") +
  labs(title = "SPEI_har_pond Smooth Terms with Different k",
       x = "SPEI_har_pond", y = "Smooth Estimate",
       color = "k value") +
  theme_minimal()

# Display the plots
p1

```
K adjustment
Median of all the k combinations that have less than 2 points more than the min AICc

Avec Dredge
```{r}

#retirer les effets randoms
model_gam <- gam(ratio ~ s(SPEI_har_pond, bs = "cr", k=3) + SEX + CHICKAGE,
                 family = binomial(link = "logit"),
                 data = test, na.action = na.fail)

Dredge <- dredge(model_gam)

Dredge

```

```{r}
#automated adjustment: k Ref =  8.690

#with k = 1

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=3) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
#k a été augmenté à sa valeur minimale telle que le modèle converge = 3
gam_model <- modok1$gam
AICc(modok1)#744.304
BIC(modok1)
BIC(modok1$lme)
summary(modok1$gam)
summary(modok1$lme)
r.squaredGLMM(modok1$lme)
k.check(modok1$gam)

#juste sex et spei
modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=3) + SEX,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#2066.364
r.squaredGLMM(modok1$lme)

#juste chick
modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=3) + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#1073.691

#juste SPEI
modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=3), random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#2399.027
r.squaredGLMM(modok1$lme)

#juste aléatoire
modok1 <- gamm(ratio ~ 1,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#2575.557
r.squaredGLMM(modok1$lme)

#Juste sex
modok1 <- gamm(ratio ~ SEX,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#2285.422
r.squaredGLMM(modok1$lme)

#juste chick age
modok1 <- gamm(ratio ~ CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)
r.squaredGLMM(modok1$lme)

#SEX + CHICK AGE
modok1 <- gamm(ratio ~ CHICKAGE+ SEX,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)
r.squaredGLMM(modok1$lme)

#best = complet
modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=1) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)
r.squaredGLMM(modok1$lme)

#try other values of k
modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=2) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)# 744.304

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=3) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)# 744.304

###EGALITE car au minimum k = 3

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=4) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#744.129

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=5) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)# 738.8517

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=6) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)# 707.3365

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=7) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#732.6863

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=8) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#  660.3094 best model
BIC(modok1)
BIC(modok1$lme)
summary(modok1$gam)
r.squaredGLMM(modok1$lme)
k.check(modok1$gam)

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=9) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)# 686.8168

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=10) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#666.3258

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=11) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#645.4085

modok1 <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=12) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), data = test)
AICc(modok1)#614.0236
```
Best k = 1 (more parsimonious)
After 9 overfitting

### GAMM k1 Model validation

```{r}
modok <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=3) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), method = "REML", data = test)
AICc(modok)# 744.304
coef(modok$gam)
      #  (Intercept)               SEXM          CHICKAGE1 s(SPEI_har_pond).1 s(SPEI_har_pond).2 
      # -0.392933757       -0.301250902        0.555548341       -0.004789189       -0.150514592 
summary(modok$gam)
summary(modok$gam)$p.table
summary(modok$gam)$s.table
VarCorr(modok$lme)
summary(modok$lme)

# Variances extraites
var_Xr <- 3.667438e-03
var_POP <- 4.368242e-10
var_PRODUCTION_envol <- 2.454076e-02 
var_residual <- 1.000000e+00

# Variance totale (somme des variances)
var_total <- var_Xr + var_POP + var_PRODUCTION_envol + var_residual

# Proportions de variance expliquée
prop_Xr <- var_Xr / var_total *100
prop_POP <- var_POP / var_total*100
prop_PRODUCTION_envol <- var_PRODUCTION_envol / var_total*100
prop_residual <- var_residual / var_total*100

# Affichage des proportions
prop_Xr
prop_POP
prop_PRODUCTION_envol
prop_residual
plot(modok$gam)

#residuals
residu <- residuals(modok$gam, type = "pearson")
fitted <- fitted(modok$gam)
qqnorm(residu)
qqline(residu)
#normalité des résidus
hist(residu)

modok_gam <- modok$gam
# vif(modok_gam)
#(Zuur et al: non-linear relationships are not picked up by the VIF

# residuals vs fitted (homoscedasticity)
plot(residu~fitted
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residu~ test$CHICKAGE,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Chick age",
         main = "")
abline(h = 0)

plot(residu~ test$SEX,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SEX",
         main = "")
abline(h = 0)

plot(residu~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)


#R²
summary(modok$gam)#R-sq.(adj) =   0.6  
summary(modok$lme)
VarCorr(modok$lme)
#BIC
BIC(modok$lme)# 642.0765
BIC(modok) #
#AICc
AICc(modok)


#Avons-nous optimisé le compromis entre le lissage (λ) et la flexibilité (k) dans notre modèle?
#is k sufficient?
gam.check(modok_gam)
#p-value >0.05 = yes k' ne doit pas être trop proche de edf

# k’ = 2.00, edf = 1.88 : La dimension effective (edf) est proche de k’, ce qui suggère que le modèle utilise presque toute la flexibilité disponible.
# 
# k-index = 0.84 : L’indice est inférieur à 1 mais pas extrêmement faible. Cela peut indiquer une légère contrainte due à une dimension de base insuffisante.
# 
# p-value = 0.14 : Une valeur p > 0.05 suggère qu'il n'y a pas de preuve statistique que k’ est trop faible.

#IConfidence interval (95%) for each variable
confint(modok$gam, oldNames=FALSE)
plot(modok$gam) #hard to interpret, is k really ok and not too high?
summary(modok$lme)

### Prédictions pour avoir un estimate du SPEI
# Calculer les prédictions du modèle
predictions <- predict(modok_gam$gam, type = "terms", se.fit = TRUE)

# Extraire l'estimation pour le terme smoothé 's(SPEI_har_pond)'
smooth_estimate <- predictions$fit[, "s(SPEI_har_pond)"]

# Si vous voulez aussi obtenir les erreurs standards associées à ces prédictions
se_smooth_estimate <- predictions$se.fit[, "s(SPEI_har_pond)"]

# Afficher les estimations
head(smooth_estimate)
head(se_smooth_estimate)


#autocorrélation temporelle eds résidus ?
par(mfrow = c(1,2))
acf(resid(modok_gam), lag.max = 36, main = "ACF")
pacf(resid(modok_gam), lag.max = 36, main = "pACF")
#2 et 1 pic seulemnt

plot(modok_gam, select = 1)

test$predicted <- predict(modok, type = "response")

ggplot(test, aes(x = SPEI_har_pond, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "blue") +
  labs(title = "GAMM Predictions",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()



#############

# Créer une grille de nouvelles données pour prédire les effets
new_data <- expand.grid(SPEI_har_pond = seq(min(test$SPEI_har_pond), max(test$SPEI_har_pond), length.out = 100),
                        SEX = factor(c('M', 'F'), levels = levels(test$SEX)),
                        CHICKAGE = factor(c('1', '2'), levels = levels(test$CHICKAGE)))

# Prédire avec le modèle
preds <- predict(modok$gam, newdata = new_data, type = "link", se.fit = TRUE)

# Ajouter les prédictions à la grille de nouvelles données
new_data$fit <- preds$fit
new_data$se.fit <- preds$se.fit

# Calculer les intervalles de confiance à 95% sur l'échelle du lien
new_data$upper <- new_data$fit + 1.96 * new_data$se.fit
new_data$lower <- new_data$fit - 1.96 * new_data$se.fit

# Convertir les intervalles de confiance à l'échelle de la réponse
new_data$pred <- modok$gam$family$linkinv(new_data$fit)
new_data$pred_upper <- modok$gam$family$linkinv(new_data$upper)
new_data$pred_lower <- modok$gam$family$linkinv(new_data$lower)

ggplot(new_data, aes(x = SPEI_har_pond, y = pred, color = SEX, linetype = CHICKAGE)) +
    geom_line() +
    geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper, fill = SEX), alpha = 0.2) +
    labs(title = "Effet de SPEI_har_pond sur ratio par SEX et CHICKAGE",
         x = "SPEI_har_pond", y = "Prédiction du ratio") +
    theme_minimal() +
    scale_fill_manual(values = c('blue', 'red')) +  # Ajuster les couleurs si nécessaire
    scale_color_manual(values = c('blue', 'red'))  # Ajuster les couleurs si nécessaire

##################

# Créer une nouvelle grille de valeurs pour SPEI_har_pond
new_data <- data.frame(SPEI_har_pond = seq(min(test$SPEI_har_pond), max(test$SPEI_har_pond), length.out = 100),
                       SEX = factor("F"),  # Fixer les niveaux de SEX et CHICKAGE pour simplifier
                       CHICKAGE = factor("1"))

# Prédire les valeurs de réponse et les intervalles de confiance
preds <- predict(modok$gam, newdata = new_data, se.fit = TRUE, type = "link")

# Calculer les intervalles de confiance à 95%
new_data$fit <- preds$fit
new_data$se.fit <- preds$se.fit
new_data$upper <- preds$fit + 1.96 * preds$se.fit
new_data$lower <- preds$fit - 1.96 * preds$se.fit
new_data$pred <- modok$gam$family$linkinv(preds$fit)
new_data$pred_upper <- modok$gam$family$linkinv(new_data$upper)
new_data$pred_lower <- modok$gam$family$linkinv(new_data$lower)


ggplot(new_data, aes(x = SPEI_har_pond, y = pred)) +
    geom_line() +
    geom_ribbon(aes(ymin = pred_lower, ymax = pred_upper), alpha = 0.2) +
    labs(title = "Effet de SPEI_har_pond sur ratio",
         x = "SPEI_har_pond", y = "Prédiction du ratio") +
    theme_minimal()



p_obj <- plot(modok$gam, residuals = TRUE)
p_obj <- p_obj[[1]] # just one smooth so select the first component
sm_df <- as.data.frame(p_obj[c("x", "se", "fit")])
data_df <- as.data.frame(p_obj[c("raw", "p.resid")])

## plot
ggplot(sm_df, aes(x = x, y = fit)) +
  geom_rug(data = data_df, mapping = aes(x = raw, y = NULL),
           sides = "b") +
  geom_point(data = data_df, mapping = aes(x = raw, y = p.resid)) +
  geom_ribbon(aes(ymin = fit - se, ymax = fit + se, y = NULL),
              alpha = 0.3) +
  geom_line() +
  labs(x = p_obj$xlab, y = p_obj$ylab)
```

#représentation2

```{r}
modok <- gamm(ratio ~ s(SPEI_har_pond, bs = "cr", k=1) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), method = "REML", data = test)


newdat <- expand.grid(SPEI_har_pond = seq(-1.469, 1.76, 0.1),
                      SEX = unique(test$SEX),
                      CHICKAGE = unique(test$CHICKAGE))
library(stats)
newdat$logit_pred <- predict(modok, newdata = newdat, type = "link", se.fit = TRUE)$fit
newdat$logit_SE <- predict(modok, newdata = newdat, type = "link", se.fit = TRUE)$se.fit
cmult <- 1.96  # Pour un intervalle de confiance de 95%
newdat$logit_SE_lower <- newdat$logit_pred - cmult * newdat$logit_SE
newdat$logit_SE_upper <- newdat$logit_pred + cmult * newdat$logit_SE

# Transformer les prédictions et les intervalles de confiance en probabilités
newdat$prob_pred <- plogis(newdat$logit_pred)
newdat$prob_SE_lower <- plogis(newdat$logit_SE_lower)
newdat$prob_SE_upper <- plogis(newdat$logit_SE_upper)


g0 <- ggplot(newdat, aes(x = SPEI_har_pond, y = prob_pred, colour = SEX)) + 
  geom_line() +
  geom_point(data = test, aes(x = SPEI_har_pond, y = prop_foraging, col = SEX)) +
  geom_ribbon(aes(ymin = prob_SE_lower, ymax = prob_SE_upper, fill = SEX), alpha = 0.1) +
  facet_wrap(~ CHICKAGE, scales = "fixed", ncol = 2) +
  scale_color_manual(values = c("darkred", "deepskyblue4")) +  # Les lignes noires pour les deux groupes
  scale_fill_manual(values = c("M" = "deepskyblue", "F" = "red")) + # Facetter par CHICKAGE avec échelles Y fixes
  theme_classic() +
  labs(x = "SPEI", y = "Foraging proportion per month") +  # Adapté à votre modèle
  theme(
    axis.text.x = element_text(size = 16),  # Increase size of x-axis text
    axis.text.y = element_text(size = 16),  # Increase size of y-axis text
    axis.title.x = element_text(size = 18), # Increase size of x-axis title
    axis.title.y = element_text(size = 18), # Increase size of y-axis title
    strip.text = element_blank(),
    legend.position = "none" # Remove facet labels
  )

```

#Test sans le smois incomplets
```{r}
id_30_to_delete <- c("190657_2021_2", "190665_2020_1", "200015_2020_1")
test1 <-  test[!(test$id_30 %in% id_30_to_delete), ]
ratio1 <- ratio[-c(12,31,45),]

modok <- gamm(ratio1 ~ s(SPEI_har_pond, bs = "cr", k=3) + SEX + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1),family = binomial(link = "logit"), method = "REML", data = test1)
AICc(modok)# 750.1146
coef(modok$gam)
 # (Intercept)               SEXM          CHICKAGE1 s(SPEI_har_pond).1 s(SPEI_har_pond).2 
 #      -0.385952979       -0.305170152        0.554848066       -0.002099001       -0.147879233 
summary(modok$gam)
summary(modok$gam)$p.table
summary(modok$gam)$s.table
VarCorr(modok$lme)
summary(modok$lme)
```


##GLMM
```{r}
#SPEI
modglm <- glmer(ratio~  SPEI_har_pond + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), family = binomial(link = "logit"), data = test, na.action = na.fail)
AICc(modglm)#1297.616

#Re-check the factors
#whithout SEX
modglm1 <- glmer(ratio~    SPEI_har_pond +  CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), family = binomial(link = "logit"),data = test, na.action = na.fail)
AICc(modglm1) #  1305.035

#whithout CHICKAGE
modglm2 <- glmer(ratio~  SPEI_har_pond +SEX +
                  (1|POP/device) + (1|PRODUCTION_envol), family = binomial(link = "logit"),data = test, na.action = na.fail)
AICc(modglm2) #2660.24 keep them both

#  Les résidus ne sont pas surdispersés ?
# que leur variance n’est pas plus grande que celle admise par le modèle ?
#overdisp.glmer(modglm2) #surdispersé

########################

#SPEI²
test$SPEI_har_sq <- test$SPEI_har_pond^2

modglm <- glmer(ratio~ SPEI_har_sq + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)
AICc(modglm) #1365.075 not better

########################

#SPEI AND SPEI² (polynomial function)
modglm <- glmer(ratio~ SPEI_har_sq + SPEI_har_pond + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)
AICc(modglm) #1297.281 not better

#Standardized residuals vs predicted values: if random distribution = ok
plot(modglm, which = 1) #ok

```

### Model validation: assumption checking

```{r}
#selected model
#REML = T for model validation (provides less biased estimates of the variance components)
modok <- glmer(ratio~ SPEI_har_pond+ I(SPEI_har_pond^3) + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)
AICc(modok)#1294.719

#residuals (normality and independance)
qqnorm(resid(modok))
qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test$CHICKAGE,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Chick age",
         main = "")
abline(h = 0)

plot(residuals(modok)~ test$SEX,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SEX",
         main = "")
abline(h = 0)


plot(residuals(modok)~ test$SPEI_har_pond,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SPEI_har_pond",
         main = "")
abline(h = 0)


plot(Effect("SPEI_har_pond",modok,partial.residuals=TRUE))

#Multicolinearity
vif(modok)

#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#                   R2m      R2c
# theoretical 0.6417333 0.979656
# delta       0.6386309 0.974920

BIC(modok)

#IConfidence interval (95%) for each variable
# confint(modok, oldNames=FALSE)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

summary(modok)
#               Estimate Std. Error z value Pr(>|z|)    
# (Intercept)   -0.32534    0.09185  -3.542 0.000397 ***
# SPEI_har_pond -0.06664    0.01037  -6.426 1.31e-10 ***
# SEXM          -0.33733    0.09199  -3.667 0.000245 ***
# CHICKAGE2      0.52366    0.01553  33.718  < 2e-16 ***



###################################

#selected model
#REML = T for model validation (provides less biased estimates of the variance components)
modok <- glmer(ratio~ SPEI_har_sq + SPEI_har_pond + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)

#residuals (normality and independance)
qqnorm(resid(modok))
qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test$CHICKAGE,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Chick age",
         main = "")
abline(h = 0)

plot(residuals(modok)~ test$SEX,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SEX",
         main = "")
abline(h = 0)


plot(residuals(modok)~ test$SPEI_har_pond,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SPEI_har_pond",
         main = "")
abline(h = 0)

plot(residuals(modok)~ test$SPEI_har_sq,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SPEI_har_sq",
         main = "")
abline(h = 0)


plot(Effect("SPEI_har_pond",modok,partial.residuals=TRUE))
plot(Effect("SPEI_har_sq",modok,partial.residuals=TRUE))

#Multicolinearity
vif(modok)
  # SPEI_har_sq SPEI_har_pond           SEX      CHICKAGE 
  #    1.021550      1.104682      1.001068      1.082415 
     
#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#                   R2m       R2c
# theoretical 0.6538248 0.9808269
# delta       0.6508451 0.9763569

#IConfidence interval (95%) for each variable
# confint(modok, oldNames=FALSE)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SPEI_har_sq + SEX + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

lm_resid <- lm(residus^2 ~ SPEI_har_sq + SEX + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

summary(modok)
#               Estimate Std. Error z value Pr(>|z|)    
# (Intercept)   -0.39783    0.09512  -4.183 2.88e-05 ***
# SPEI_har_sq    0.07912    0.00948   8.346  < 2e-16 ***
# SPEI_har_pond -0.07913    0.01047  -7.559 4.05e-14 ***
# SEXM          -0.32460    0.09864  -3.291 0.000999 ***
# CHICKAGE2      0.51969    0.01554  33.431  < 2e-16 ***

```





# Test relation quadratique
```{r}
#With the quadratic variable
# The I() function ensures that the enclosed mathematical operation 
# is correctly interpreted as part of the model specification
modok <- glmer(ratio~ SPEI_har_pond + I(SPEI_har_pond^2) + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)

AICc(modok) #1297.281
r.squaredGLMM(modok)
#                   R2m       R2c
# theoretical 0.6538315 0.9808270
# delta       0.6508517 0.9763571
vif(modok)
    # SPEI_har_pond I(SPEI_har_pond^2)                SEX           CHICKAGE 
    #       1.104682           1.021550           1.001068           1.082415 
plot(Effect("SPEI_har_pond",modok,partial.residuals=TRUE))

library(sjPlot)
plot_model(modok, type = "pred", terms = "SPEI_har_pond[all]", ci.lvl = 0.95) +
  labs(title = "Polynomial Effect of SPEI_har_pond on Ratio",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()


# Extract model predictions
test$predicted <- predict(modok, type = "response")


# Plot using ggplot2
ggplot(test, aes(x = SPEI_har_pond, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE, color = "blue") +
  labs(title = "Polynomial Effect of SPEI_har_pond on Ratio",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()
```

#######################

#Test cubic variable

```{r}
#With the quadratic variable
# The I() function ensures that the enclosed mathematical operation 
# is correctly interpreted as part of the model specification
modok <- glmer(ratio~ SPEI_har_pond + I(SPEI_har_pond^2) + I(SPEI_har_pond^3)  + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)

AICc(modok) #1296.299
plot(Effect("SPEI_har_pond",modok,partial.residuals=TRUE))

plot_model(modok, type = "pred", terms = "SPEI_har_pond[all]", ci.lvl = 0.95) +
  labs(title = "Polynomial Effect of SPEI_har_pond on Ratio",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()


# Extract model predictions
test$predicted <- predict(modok, type = "response")


# Plot using ggplot2
ggplot(test, aes(x = SPEI_har_pond, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), se = TRUE, color = "blue") +
  labs(title = "Polynomial Effect of SPEI_har_pond on Ratio",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()
#################
#terme cubique seul avec SPEI ?

modok <- glmer(ratio~ SPEI_har_pond+ I(SPEI_har_pond^3)  + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)
AICc(modok)#1294.719
BIC(modok)

```
#Test quartic

```{r}
#With the quadratic variable
# The I() function ensures that the enclosed mathematical operation 
# is correctly interpreted as part of the model specification
modok <- glmer(ratio~ SPEI_har_pond + I(SPEI_har_pond^2) + I(SPEI_har_pond^3) + I(SPEI_har_pond^4)  + SEX + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol),family = binomial(link = "logit"), data = test, na.action = na.fail)

AICc(modok) #1237.239
plot(Effect("SPEI_har_pond",modok,partial.residuals=TRUE, confidence.level = 0.95))

plot_model(modok, type = "pred", terms = "SPEI_har_pond[all]", ci.lvl = 0.95) +
  labs(title = "Polynomial Effect of SPEI_har_pond on Ratio",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()


# Extract model predictions
test$predicted <- predict(modok, type = "response")


# Plot using ggplot2
ggplot(test, aes(x = SPEI_har_pond, y = predicted)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 4), se = TRUE, color = "blue") +
  labs(title = "Polynomial Effect of SPEI_har_pond on Ratio",
       x = "SPEI_har_pond",
       y = "Predicted Ratio") +
  theme_minimal()
#################
```





