---
title: "mean_dist_nest_model"
output:
  html_document: default
  pdf_document: default
date: "2024-05-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim of this script is to statistically test the link between average daily distance from the nest and various factors as explanatory variables (drought index, sex, age of chicks, etc.) via the construction of a mixed model.

## Load packages

```{r, message= FALSE, warning=FALSE}
library(dplyr)
library(corrplot)
library(lme4) # for lmer
library(nlme) #for gls (glmm)
library(lmerTest)
library(Matrix) 
library(MuMIn) # for Dredge()
library(car) # for vif()
library(lubridate)
library(GGally) #correlation synthese
library(mgcv) #for gam
library(lmtest) #Breusch-Pagan test
library(effects)
library(flexmix) #for BIC extraction
library(performance) #R²
library(gridExtra)#grid.arrange
```

## Prepare DATA

```{r, warning=FALSE, message=FALSE}
data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025.txt",
    header = T,
    dec = ","
  )
#79 periods

data <-
  data1[, c(
    "device",
    "id_30",
    "Deviceyear",
    "POP",
    "SEXE",
    "PRODUCTION_envol",
    "CHICKAGE",
    "SPEI_har_pond",
    "Mean_nest_dist"
  )]

#Convert into the good type
data <- data %>%
  mutate(across(c(POP, SEXE, PRODUCTION_envol, CHICKAGE), as.factor)) %>%
  mutate(across(c(SPEI_har_pond, Mean_nest_dist), as.numeric))


# With the default settings, the scale() function calculates the vector’s mean and standard deviation, here it is not needed as we have only one numeric explictaive variable
# then “scales” each element by removing the mean and dividing by the sd.
#Here it is necessary sinc the meteorological descriptors have different units

table(data$PRODUCTION_envol)
# Only one individual has 5 chicks :add it to the 4+ class
data$PRODUCTION_envol <-
  ifelse(
    data$PRODUCTION_envol == 4 |
      data$PRODUCTION_envol == 5,
    "4+",
    as.character(data$PRODUCTION_envol)
  )

data$PRODUCTION_envol <- as.factor(data$PRODUCTION_envol)
data$device <- as.factor(data$device)
```

## Explicative variables description and distribution

```{r}

#SPEI Hargreaves
dotchart(data$SPEI_har_pond,pch=16,col='blue',xlab='SPEI')
# Histogram
hist(data$SPEI_har_pond,col='blue',xlab="SPEI",main="") #not normal

# Quantile-Quantile plot
qqnorm(data$SPEI_har_pond,pch=16,col='blue',xlab='')
qqline(data$SPEI_har_pond,col='red') #not normal

#factors
#Chick production 
table(data$PRODUCTION_envol) #pas équilibré

#SEXE
table(data$SEX)

#Chick age
table(data$CHICKAGE)
```

#Y distribution

```{r}
test <-
  data[, c(
    "device",
    "id_30",
    "Mean_nest_dist",
    "POP",
    "SEXE",
    "PRODUCTION_envol",
    "CHICKAGE",
    "SPEI_har_pond"
  )]
test$PRODUCTION_envol <- as.factor(test$PRODUCTION_envol)
test$device <- as.factor(test$device)

#Check for missing values
colSums(is.na(test))

#Y distribution

par(mfrow = c(2, 2))
# Boxplot
boxplot(test$Mean_nest_dist, col = 'blue', ylab = 'Mean daily distances to the nest')
# Cleveland plot
dotchart(
  test$Mean_nest_dist,
  pch = 16,
  col = 'blue',
  xlab = 'Mean daily distances to the nest'
)
# Histogram
hist(
  test$Mean_nest_dist,
  col = 'blue',
  xlab = "Mean daily distances to the nest",
  main = ""
)
# Quantile-Quantile plot
qqnorm(
  test$Mean_nest_dist,
  pch = 16,
  col = 'blue',
  xlab = ''
)
qqline(test$Mean_nest_dist, col = 'red')
# Non normal
shapiro.test(test$Mean_nest_dist) #W = 0.80189, p-value = 6.279e-09 not normal
#No outlier
```

No outliers, non-linear distribution

### Potential relationships between Xs and Y

```{r}
par(mfrow=c(1,4))

# SPEI
plot(
  test$Mean_nest_dist ~ test$SPEI_har_pond,
  pch = 16,
  col = 'blue',
  xlab = 'SPEI',
  ylab = 'Mean distance to the nest'
)

# Production
boxplot(Mean_nest_dist ~ PRODUCTION_envol,
        data = test,
        main = "Mean distance to the nest")
# sex
boxplot(
  Mean_nest_dist ~ SEXE,
  data = test,
  col = c('blue3', 'red2'),
  main = "Mean distance to the nest"
)
# Chick age
boxplot(
  Mean_nest_dist ~ CHICKAGE,
  data = test,
  col = c('blue3', 'red2'),
  main = "Mean distance to the nest"
)
```

Mean distance to the nest: decreases with the number of chicks, increases in males, slightly increases with the chick age (not tested yet)

### Choice of the random factors

Stepwise forward selection beginning with a LMM The best random structure minimizes the AICc and is the most parsimonious with a delta AICc \< 2 (if several models AICc are close from each other and minimum)

```{r}
#LM, constant model
mod <- lm(Mean_nest_dist~1, data = test)
summary(mod)

#LMM
mod1 <- lmer(Mean_nest_dist~(1|POP/device), data = test)
#summary(mod1)
mod2 <- lmer(Mean_nest_dist~(1|POP/device) + (1|PRODUCTION_envol), data = test)
#summary(mod2)
mod3 <- lmer(Mean_nest_dist~(1|PRODUCTION_envol), data = test)
#summary(mod3)
mod4 <- lmer(Mean_nest_dist~(1|device), data = test)
#summary(mod4)
mod5 <- lmer(Mean_nest_dist~(1|device) + (1|POP), data = test)
#summary(mod5)
mod6 <- lmer(Mean_nest_dist~ (1|POP), data = test)
#summary(mod6)

#Compare AICc
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6)
aiccs <- sapply(model_list, AICc)
aiccs #1403.883 1379.936 1376.009 1389.023 1378.195 1379.936 1389.694
best_model <- model_list[[ which.min(aiccs)]]
best_model 
#Mean_nest_dist ~ (1 | POP/device) + (1 | PRODUCTION_envol) 1376.009 and delta AICc >2
```

\~ (1 \| POP/device) + (1 \| PRODUCTION_envol) is the selected random structure

### GAMM with the complete model: linearity of the relationship?

The Edf (Effective degree of Freedom) indicates the degree of complexity for each relationship \n

The Bayesian model of spline smoothing introduced by Wahba (1983) and Silverman (1985) opens up the possibility of estimating the degree of smoothness of terms in a generalized additive model as variances of the wiggly components of the smooth terms treated as random effects The default method is REML (Restricted Maximum Likelihood), it generally provides better estimates of variance components and is less biased, which is advantageous in most applications involving mixed-effects models. LM : maximum likelihood, is better to compare fixed effects combinations

```{r, warning=FALSE , message=FALSE}

#without any factor
modgam <- gamm(Mean_nest_dist~s(SPEI_har_pond),  random = list(POP = ~1|device, PRODUCTION_envol = ~1), method = "ML", data = test)
AICc(modgam) # 1383.026


#remove SEX
modgam <- gamm(Mean_nest_dist~ s(SPEI_har_pond) + 
                 CHICKAGE, random = list(POP = ~1|device, PRODUCTION_envol = ~1), method = "ML", data = test)
AICc(modgam)# 1383.922 delta <2 it doesn't improve the model so remove it

#remove chick age
modgam <- gamm(Mean_nest_dist~s(SPEI_har_pond) + 
                 SEXE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1), method = "ML", data = test)
AICc(modgam)# 1384.386 idem

#avec les deux
modgam <- gamm(Mean_nest_dist~s(SPEI_har_pond) + 
                 SEXE + CHICKAGE,  random = list(POP = ~1|device, PRODUCTION_envol = ~1), method = "ML", data = test)
AICc(modgam)#1385.374

summary(modgam$gam) #Edf = 1 for SPEI
#Standardized residuals vs predicted values: random distribution = ok (exploration here)
plot(modgam$lme, which = 1) # heteroscedasticity, 
#will probably need a transformation -> see with the GLMM model
```

Edf = 1 --\> linear relationship --\> LMM

## LMM model backward selection

To select the significant explicative variables

REML = F (so ML) ML ensures that the likelihood function is calculated in a consistent manner across different models, making AIC (Akaike Information Criterion) or AICc (corrected AIC) comparisons valid.

```{r}
#selected model
modglm <- lmer(Mean_nest_dist~  SPEI_har_pond + SEXE + CHICKAGE +
                  (1 | POP/device) + (1 | PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm)#1389.625

#whithout SEXE
modglm1 <- lmer(Mean_nest_dist~ SPEI_har_pond +  CHICKAGE +
                 (1 | POP/device) + (1 | PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm1) # 1388.924, delta aic <2, remove it

#whithout CHICKAGE
modglm2 <- lmer(Mean_nest_dist~SPEI_har_pond + SEXE +
                 (1 | POP/device) + (1 | PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm2) # 1388.978 delta <2
#keep the most parsimonious

#only SPEI
modglm4 <- lmer(Mean_nest_dist~  SPEI_har_pond +
                 (1 | POP/device) + (1 | PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm4)# 1388.437


#Standardized residuals vs predicted values: if random distribution = ok
plot(modglm4, which = 1) #Heteroscedasticity ! Need a transformation
```

Singular pattern --\> heteroskedasticity !

### #Log transformation to avoid heteroskedasticity:

```{r}
test$Mean_nest_dist_log <- log(test$Mean_nest_dist)

#new distribution
par(mfrow = c(2, 2))
# Boxplot
boxplot(test$Mean_nest_dist_log,
        col = 'blue',
        ylab = 'Mean of the monthly mean (log) of daily distances to the nest')
# Cleveland plot
dotchart(
  test$Mean_nest_dist_log,
  pch = 16,
  col = 'blue',
  xlab = 'Mean of the monthly mean (log) of daily distances to the nest'
)
# Histogram
hist(
  test$Mean_nest_dist_log,
  col = 'blue',
  xlab = "Mean of the monthly mean (log) of daily distances to the nest",
  main = ""
)
# Quantile-Quantile plot
qqnorm(
  test$Mean_nest_dist_log,
  pch = 16,
  col = 'blue',
  xlab = ''
)
qqline(test$Mean_nest_dist_log, col = 'red')

shapiro.test(test$Mean_nest_dist)# W = 0.80189, p-value = 6.279e-09 Non normal
shapiro.test(test$Mean_nest_dist_log) #W = 0.98495, p-value = 0.4809 Normal
#No outlier
```

### LOG Choice of the random factors

Same process as before

```{r}
#LM, constant model
mod <- lm(Mean_nest_dist_log~1, data = test)
AICc(mod)#202.0492

#LMM
mod1 <- lmer(Mean_nest_dist_log~(1|POP/device), data = test)
AICc(mod1)#186.631
mod2 <- lmer(Mean_nest_dist_log~(1|POP/device) + (1|PRODUCTION_envol), data = test)
AICc(mod2)#182.9778
mod3 <- lmer(Mean_nest_dist_log~(1|PRODUCTION_envol), data = test)
AICc(mod3)# 202.3324
mod4 <- lmer(Mean_nest_dist_log~(1|device), data = test)
AICc(mod4)#185.7412
mod5 <- lmer(Mean_nest_dist_log~(1|device) + (1|POP), data = test)
AICc(mod5)#186.631
mod6 <- lmer(Mean_nest_dist_log~ (1|POP), data = test)
AICc(mod6)#199.255

#Compare AICc
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6)
aiccs <- sapply(model_list, AICc)
aiccs #202.0492 186.6310 182.9778 202.3324 185.7412 186.6310 199.2550
best_model <- model_list[[ which.min(aiccs)]]
best_model 
#Mean_nest_dist ~ (1 | POP/device) + (1 | PRODUCTION_envol)  
```

It keeps the same random structure

### LOG GAMM with the complete log model: linearity of the relationship?

```{r}
#log transformation seems to be useful: complete model
modgam1 <-
  gamm(
    Mean_nest_dist_log ~ s(SPEI_har_pond) + SEXE +
      CHICKAGE,
    random = list(POP = ~ 1 |
                    device, PRODUCTION_envol = ~ 1),
    method = "ML",
    data = test
  )
AICc(modgam1) #193.8934

#Try to remove SEX
modgam2 <- gamm(
  Mean_nest_dist_log ~  s(SPEI_har_pond) +
    CHICKAGE,
  random = list(POP = ~ 1 |
                  device, PRODUCTION_envol = ~ 1),
  method = "ML",
  data = test
)
AICc(modgam2)#192.8182 delta AICC < 2 so remove it

#Try to remove Chickage
modgam3 <-
  gamm(
    Mean_nest_dist_log ~ s(SPEI_har_pond) + SEXE,
    random = list(POP = ~ 1 |
                    device, PRODUCTION_envol = ~ 1),
    method = "ML",
    data = test
  )
AICc(modgam3)# 192.1515 idem

#without the factors
modgam <-
  gamm(
    Mean_nest_dist_log ~ s(SPEI_har_pond),
    random = list(POP = ~ 1 |
                    device, PRODUCTION_envol = ~ 1),
    method = "ML",
    data = test
  )
AICc(modgam) #191.1692

plot(modgam$gam)
summary(modgam$gam) #edf = 1 -> linear relationship

# Each plot represents the smoothed effect of a continuous variable on the response variable.
# The solid line is the estimated effect, and the shaded areas represent the 95% confidence intervals.
summary(modgam$lme)
```

Linear relationship for SPEI --\> LMM

## LOG LMM model backward selection

REML = F (so ML) ML ensures that the likelihood function is calculated in a consistent manner across different models, making AIC (Akaike Information Criterion) or AICc (corrected AIC) comparisons valid.

```{r}
#complete model
modglm <- lmer(Mean_nest_dist_log~  SPEI_har_pond + SEXE + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm)#181.4757
r.squaredGLMM(modglm)

#whithout SEXE
modglm2 <- lmer(Mean_nest_dist_log~   SPEI_har_pond +  CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm2) #180.054 , delta AICc <2, remove it
r.squaredGLMM(modglm2)

#whithout CHICKAGE
modglm3 <- lmer(Mean_nest_dist_log~  SPEI_har_pond + SEXE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm3) #180.4939 idem
r.squaredGLMM(modglm3)

#Only SPEI
modglm4 <- lmer(Mean_nest_dist_log~  SPEI_har_pond +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm4) #179.2096
r.squaredGLMM(modglm4)

#only sex
modglm4 <- lmer(Mean_nest_dist_log~  SEXE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm4) #184.1427
r.squaredGLMM(modglm4)

#only chickage
modglm4 <- lmer(Mean_nest_dist_log~  CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm4) #183.4918
r.squaredGLMM(modglm4)

#only sex + chickage
modglm4 <- lmer(Mean_nest_dist_log~  CHICKAGE + SEXE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
AICc(modglm4) #184.9909
r.squaredGLMM(modglm4)

#NULL model
modglm5 <-
  lmer(
    Mean_nest_dist_log ~ (1 | POP / device) + (1 | PRODUCTION_envol),
    data = test,
    REML = F,
    na.action = na.fail
  )
AICc(modglm5) #182.7731 (delta AICc >2 so the best model is modglm4)
r.squaredGLMM(modglm5)

#selected model
modglm <-
  lmer(
    Mean_nest_dist_log ~  SPEI_har_pond + (1 | POP / device) + (1 |PRODUCTION_envol),
    data = test,
    REML = F,
    na.action = na.fail
  )
#Standardized residuals vs predicted values: random distribution = ok
plot(modglm, which = 1) #Homoscedasticity ok
#AICc(modglm)#174.2247
```

Homoskedasticity ok !

Avec Dredge
```{r}

fullmod <- lmer(Mean_nest_dist_log~  SPEI_har_pond + SEXE + CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = F, na.action = na.fail)
Dredge <- dredge(fullmod)

#3 modèles en concurrence
#Estimate nul pour ajout de SEX ou de CHICKAGE ?
modglm2 <- lmer(Mean_nest_dist_log~   SPEI_har_pond +  CHICKAGE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = T, na.action = na.fail)
confint(modglm2, oldNames=FALSE) #CHICKAGE ESTIMATE NULL
summary(modglm2)

modglm3 <- lmer(Mean_nest_dist_log~  SPEI_har_pond + SEXE +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = T, na.action = na.fail)
confint(modglm3, oldNames=FALSE) 
summary(modglm3)

summary(model.avg(Dredge, subset = delta<2))
```


## LOG Model validation: assumption checking

```{r}
#selected model
#REML = T for model validation (provides less biased estimates of the variance components)
modok <- lmer(Mean_nest_dist_log~ SPEI_har_pond +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test, REML = T, na.action = na.fail)
AICc(modok)#182.4488

#residuals (normality and independance)
qqnorm(resid(modok))

qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#             R2m       R2c
# [1,] 0.02894783 0.6969309

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEXE + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok ! p>0.005

summary(modok)

#PLot it
plot(residuals(modok)~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

```

Conclusion: - The daily mean distance to the nest decreases in wetter conditions (high SPEI) So it increases in dryer conditions

# Representation with and without the log trasnformation

```{r}
modok <- lm(Mean_nest_dist_log~ SPEI_har_pond, data = test, na.action = na.fail)

newdat <- expand.grid(SPEI_har_pond=seq(-1.469,1.76,0.1))
library(stats)
newdat$pred <- stats::predict(modok, newdat, level = 0)

resid <- residuals(modok)

# Calculer l'écart-type des résidus
sigma_lm <- sd(resid)

Designmat <- model.matrix(formula(modok)[-2], newdat)
predvar <- diag(Designmat %*% vcov(modok) %*% t(Designmat)) 
newdat$SE <- sqrt(predvar) 
newdat$SE2 <- sqrt(predvar+sigma_lm**2) #incertitude des prédiction + celle des résidus

g0 <- ggplot(newdat,aes(x=SPEI_har_pond,y=pred))+geom_line()

cmult <- 1.96  
graph1 <- g0 + geom_ribbon(aes(ymin=pred-cmult*SE,ymax=pred+cmult*SE), alpha = 0.1,linetype = 0, fill = "deepskyblue") +
  geom_point(data = test, aes(x = SPEI_har_pond, y = Mean_nest_dist_log)) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 16), 
        axis.text.y = element_text(size = 16), 
        axis.title.x = element_text(size = 18), 
        axis.title.y = element_text(size = 18))+
  labs(x = "SPEI", y = "Log(Distance from the nest)")

graph1



#sans le log

newdat <- expand.grid(SPEI_har_pond=seq(-1.469,1.76,0.1))
newdat$pred <- stats::predict(modok, newdat, level = 0)

resid <- residuals(modok)

# Calculer l'écart-type des résidus
sigma_lm <- sd(resid)

Designmat <- model.matrix(formula(modok)[-2], newdat)
predvar <- diag(Designmat %*% vcov(modok) %*% t(Designmat)) 
newdat$SE <- sqrt(predvar) 
newdat$SE2 <- sqrt(predvar+sigma_lm**2)
newdat$pred_nonlog <- exp(newdat$pred)
newdat$SE_nonlog <- exp(newdat$SE)
newdat$SE2_nonlog <- exp(newdat$SE2)

newdat$pred_exp <- exp(newdat$pred)
newdat$SE_exp_lower <- exp(newdat$pred - cmult * newdat$SE)
newdat$SE_exp_upper <- exp(newdat$pred + cmult * newdat$SE)

g0 <- ggplot(newdat,aes(x=SPEI_har_pond,y=pred_exp))+geom_line()

graph1 <- g0 +
  geom_ribbon(aes(ymin = SE_exp_lower, ymax = SE_exp_upper), fill = "deepskyblue", alpha = 0.1, linetype = 0) +
  geom_point(data = test, aes(x = SPEI_har_pond, y = exp(Mean_nest_dist_log))) +
  theme_bw() +
  labs(x = "SPEI", y = "Distance from the nest")

# Afficher le graphique
graph1

```

Test du modèle en enlevant le smois de - de 21j

```{r}
id_30_to_delete <- c("190657_2021_2", "190665_2020_1", "200015_2020_1")
test1 <-  test[!(test$id_30 %in% id_30_to_delete), ]


#selected model
#REML = T for model validation (provides less biased estimates of the variance components)
modok <- lmer(Mean_nest_dist_log~ SPEI_har_pond +
                  (1|POP/device) + (1|PRODUCTION_envol), data = test1, REML = T, na.action = na.fail)
AICc(modok)#178.942

#residuals (normality and independance)
qqnorm(resid(modok))

qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test1$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#             R2m       R2c
# [1,] 0.02894783 0.6969309 ---> 0.03287128 0.6738673

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEXE + CHICKAGE, data = test1)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok ! p>0.005

summary(modok)

#PLot it
plot(residuals(modok)~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

```


Correlation matrix between the 4 descriptors
```{r}

data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025_odba.txt",
    header = T,
    dec = ","
  )
#79 periods

summary(data1)
hist(data1$HRker95_foraging)
hist(data1$Mean_nest_dist)
hist(data1$ratio_foraging_all)
hist(data1$daily_mean_ODBA)

shapiro.test(data1$HRker95_foraging)
shapiro.test(data1$Mean_nest_dist)
shapiro.test(data1$ratio_foraging_all)
shapiro.test(data1$daily_mean_ODBA)

cor(data1[, c("HRker95_foraging", "Mean_nest_dist", "ratio_foraging_all", "daily_mean_ODBA")], method = "spearman")

library(corrplot)
# Calcul de la matrice de corrélation avec Spearman
corr_matrix <- cor(data1[, c("HRker95_foraging", "Mean_nest_dist", "ratio_foraging_all", "daily_mean_ODBA")], 
                   method = "spearman")

# Ordre personnalisé (remplacez par l'ordre souhaité)
custom_order <- c("HRker95_foraging", "Mean_nest_dist", "ratio_foraging_all", "daily_mean_ODBA")

# Réorganisation de la matrice selon cet ordre
corr_matrix <- corr_matrix[custom_order, custom_order]

corrplot(corr_matrix, method = "color", type = "upper", addCoef.col = "black", tl.col = "black", tl.srt = 45)

#to have the p values
cor_test1 <- cor.test(data1$HRker95_foraging, data1$Mean_nest_dist, method = "spearman")
print(cor_test1$p.value)

# Exemple de corrélation pour HRker95_foraging et ratio_foraging_all
cor_test2 <- cor.test(data1$HRker95_foraging, data1$ratio_foraging_all, method = "spearman")
print(cor_test2$p.value)

# Exemple de corrélation pour HRker95_foraging et daily_mean_ODBA
cor_test3 <- cor.test(data1$HRker95_foraging, data1$daily_mean_ODBA, method = "spearman")
print(cor_test3$p.value)

# Exemple de corrélation pour Mean_nest_dist et ratio_foraging_all
cor_test4 <- cor.test(data1$Mean_nest_dist, data1$ratio_foraging_all, method = "spearman")
print(cor_test4$p.value)

# Exemple de corrélation pour Mean_nest_dist et daily_mean_ODBA
cor_test5 <- cor.test(data1$Mean_nest_dist, data1$daily_mean_ODBA, method = "spearman")
print(cor_test5$p.value)

# Exemple de corrélation pour ratio_foraging_all et daily_mean_ODBA
cor_test6 <- cor.test(data1$ratio_foraging_all, data1$daily_mean_ODBA, method = "spearman")
print(cor_test6$p.value)
```

