---
title: "HRkernel"
output: html_document
date: "2024-05-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r, message= FALSE, warning=FALSE}
library(dplyr)
library(corrplot)
library(lme4) # for lmer
library(nlme) #for gls (glmm)
library(lmerTest)
library(Matrix) 
library(MuMIn) # for Dredge()
library(car) # for vif()
library(lubridate)
library(GGally) #correlation synthese
library(mgcv) #for gam
library(lmtest) #Breusch-Pagan test
library(effects)
library(flexmix) #for BIC extraction
library(performance) #R²
library(gridExtra)#grid.arrange
```

#Load data \## Prepare DATA

```{r}
data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025HR.txt",
    header = T,
    dec = ","
  )
#79 periods

colnames(data1)[colnames(data1) == "SEXE"] <- "SEX"
colnames(data1)[colnames(data1) == "Hrker_foraging_hadjust"] <- "HR_ker95ha"

data <-
  data1[, c(
    "device",
    "ID",
    "id_30",
    "Deviceyear",
    "HR_ker95ha",
    "POP",
    "SEX",
    "PRODUCTION_envol",
    "CHICKAGE",
    "SPEI_har_pond"
  )]


#Convert into the good type
data <- data %>%
  mutate(across(c(POP, SEX, PRODUCTION_envol, CHICKAGE, id_30, device), as.factor))
data$SPEI_har_pond <- as.numeric(data$SPEI_har_pond)
data$HR_ker95ha <- as.numeric(data$HR_ker95ha)

#no scaling

table(data$PRODUCTION_envol)
# Only one individual has 5 chicks :add it to the 4+ class
data$PRODUCTION_envol <-
  ifelse(
    data$PRODUCTION_envol == 4 |
      data$PRODUCTION_envol == 5,
    "4+",
    as.character(data$PRODUCTION_envol)
  )
data$PRODUCTION_envol <- as.factor(data$PRODUCTION_envol)
summary(data)
```

# HR size KER95

#How to explain the monthly size of the home range calculated with KERNELS 95%?

### Prepare data

```{r}
test <- data
#Check for missing values
colSums(is.na(test))

#Y distribution

par(mfrow=c(2,2))
# Boxplot
boxplot(test$HR_ker95ha,col='blue',ylab='Monthly home range size (Ker95)')
# Cleveland plot
dotchart(test$HR_ker95ha,pch=16,col='blue',xlab='Monthly home range size (Ker95)')
# Histogram
hist(test$HR_ker95ha,col='blue',xlab="Monthly home range size (Ker95)",main="")
# Quantile-Quantile plot
qqnorm(test$HR_ker95ha,pch=16,col='blue',xlab='')
qqline(test$HR_ker95ha,col='red')
# Non normal
shapiro.test(test$HR_ker95ha) #W = 0.58959, p-value = 1.588e-13 Non normal law (chi2 ? Gamma ?)
##Outlier
```

### Potential relationships between Xs and Y

```{r}
par(mfrow=c(1,4))

# SPEI
plot(test$HR_ker95ha~test$SPEI_har_pond,pch=16,col='blue',xlab='SPEI',ylab='HR95 Kernel')

# Production
boxplot(HR_ker95ha ~ PRODUCTION_envol, data = test, main = "HR95 Kernel")
# sex
boxplot(HR_ker95ha ~ SEX, data = test, col = c('blue3','red2'), main = "HR95 Kernel")
# Chick age
boxplot(HR_ker95ha ~ CHICKAGE, data = test, col = c('blue3','red2'), main = "HR95 Kernel")
```

Outlier ! Monthly foraging HR area (ha) KER95: decreases with the number of chicks, increases in males, increases with the chick age

### Choice of the random factors

```{r}
#LM, constant model
mod <- lm(HR_ker95ha ~1, data = test)
AICc(mod)
summary(mod)

#LMM
mod1 <- lmer(HR_ker95ha ~(1|POP/device), data = test)
AICc(mod1)# 
mod2 <- lmer(HR_ker95ha ~(1|POP/device) + (1|PRODUCTION_envol), data = test)
AICc(mod2) ## !
mod3 <- lmer(HR_ker95ha ~(1|PRODUCTION_envol), data = test)
AICc(mod3)
mod4 <- lmer(HR_ker95ha ~(1|device), data = test)
AICc(mod4)
mod5 <- lmer(HR_ker95ha ~(1|device) + (1|POP), data = test)
AICc(mod5)# 
mod6 <- lmer(HR_ker95ha ~ (1|POP), data = test)
AICc(mod6)# 
mod7 <- lmer(HR_ker95ha ~ (1|PRODUCTION_envol) +(1|device), data = test)
AICc(mod7)#1521.429

#Compare AICc
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6, mod7)
aiccs <- sapply(model_list, AICc)
aiccs
best_model <- model_list[[which.min(aiccs)]] 
best_model #(1 | PRODUCTION_envol) et (1 | device) égalité
#Compare BIC
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6)
bics <- sapply(model_list, BIC)
bics
best_model <- model_list[[which.min(bics)]]
best_model 
```

### GAMM with the complete model: linearity of the relationship?

```{r}
#remove cumul and SPI (colinear variables)
modgam <- gamm(HR_ker95ha~ s(SPEI_har_pond) + SEX + 
                 CHICKAGE, random = list(device = ~1), data = test)
summary(modgam$gam)
AICc(modgam)#  1538.24

#remove SEX
modgam <- gamm(HR_ker95ha~  s(SPEI_har_pond) + CHICKAGE, random = list(device = ~1), data = test)
AICc(modgam) # 1536.883 ok remove

#remove chick age
modgam <- gamm(HR_ker95ha~  s(SPEI_har_pond) + SEX, random = list(device = ~1), data = test)
AICc(modgam) #1538.428 remove

#only SPEI
modgam <- gamm(HR_ker95ha~  s(SPEI_har_pond), random = list(device = ~1), data = test)
AICc(modgam) #1536.939 ok

#only sex
modgam <- gamm(HR_ker95ha~ SEX, random = list(device = ~1), data = test)
AICc(modgam)# 1535.968 ok

#only chick age
modgam <- gamm(HR_ker95ha~ CHICKAGE, random = list(device = ~1), data = test)
AICc(modgam)#1534.304 ok

#null model
modgam <- gamm(HR_ker95ha~ 1, random = list(device = ~1), data = test)
AICc(modgam)#1534.818 best

####With Dredge to see if we find the same model
modUgam <- uGamm(HR_ker95ha~ s(SPEI_har_pond) + SEX + 
                 CHICKAGE, random = list(device = ~1), data = test)
Dredge <- dredge(modUgam, trace = 2)
Dredge
modavg <- model.avg(Dredge, subset = delta<2)
summary(modavg)
#keep nothing then only SPEI

modgam <- gamm(HR_ker95ha~  s(SPEI_har_pond) + SEX + 
                 CHICKAGE, random = list(device = ~1), data = test)
summary(modgam$gam)
par(mfrow =c(2,2))
plot(modgam$gam)

 # Each plot represents the smoothed effect of a continuous variable on the response variable. 
 # The solid line is the estimated effect, and the shaded areas represent the 95% confidence intervals.
summary(modgam$lme)

```

Linear relationship for SPEI --\> LMM

### LMM model: selection 

```{r,results='hide',fig.keep='all'}
modglm <- lmer(HR_ker95ha~  SPEI_har_pond + SEX + CHICKAGE +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #1535.829

#remove SEX
modglm <- lmer(HR_ker95ha~  SPEI_har_pond+ CHICKAGE +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)# 1534.538 better

#remove chick age
modglm <- lmer(HR_ker95ha~  SPEI_har_pond + SEX +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#1536.113 <2 so remove it

#only SPEI
modglm <- lmer(HR_ker95ha~  SPEI_har_pond +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#1534.933 best

#only chick age
modglm <- lmer(HR_ker95ha~  CHICKAGE +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#1534.304 equal best

#only sex
modglm <- lmer(HR_ker95ha~  SEX +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#1535.968 equal best

#NUL model
modglm <- lmer(HR_ker95ha~  1 +
                  (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#1534.818 more parcimonious

modglm <- lmer(HR_ker95ha~  SPEI_har_pond + SEX + CHICKAGE +
                  (1|device), data = test, REML = F, na.action = na.fail)
```

### Model validation

```{r}

modok<- lmer(HR_ker95ha~  1+
                 (1|device), data = test, REML = T, na.action = na.fail)
#residuals (normality and independance)
qqnorm(resid(modok))
qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#Heteroscedasticity !


#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#     R2m       R2c
# [1,]   0 0.1066168

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

BIC(modok)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

summary(modok)
```

# LOG TRANSFORMATION

### LOG Y distribution

```{r}

#log transformation needed
test$HR_ker95ha_log <- log(test$HR_ker95ha)

par(mfrow=c(2,2))
# Boxplot
boxplot(test$HR_ker95ha_log,col='blue',ylab='Monthly home range size (Ker95)')
# Cleveland plot
dotchart(test$HR_ker95ha_log,pch=16,col='blue',xlab='Monthly home range size (Ker95)')
# Histogram
hist(test$HR_ker95ha_log,col='blue',xlab="Monthly home range size (Ker95)",main="")
# Quantile-Quantile plot
qqnorm(test$HR_ker95ha_log,pch=16,col='blue',xlab='')
qqline(test$HR_ker95ha_log,col='red')
# Non normal
shapiro.test(test$HR_ker95ha_log)#W = 0.97224, p-value = 0.082437 normal distribution
##it has removed the outlier
```

### LOG Potential relationships between Xs and Y

```{r}
par(mfrow=c(1,4))

# SPEI
plot(test$HR_ker95ha_log~test$SPEI_har_pond,pch=16,col='blue',xlab='SPEI',ylab='HR95 Kernel')

# Production
boxplot(HR_ker95ha_log ~ PRODUCTION_envol, data = test, main = "HR95 Kernel")
# sex
boxplot(HR_ker95ha_log ~ SEX, data = test, col = c('blue3','red2'), main = "HR95 Kernel")
# Chick age
boxplot(HR_ker95ha_log ~ CHICKAGE, data = test, col = c('blue3','red2'), main = "HR95 Kernel")
```

Monthly foraging HR area (ha) KER95: decreases with the number of chicks, increases in males, increases with the chick age ?

### LOG Choice of the random factors

```{r}
#LM, constant model
mod <- lm(HR_ker95ha_log ~1, data = test)
AICc(mod) #284.4093

#LMM
mod1 <- lmer(HR_ker95ha_log ~(1|POP/device), data = test)
AICc(mod1)#270.5984
mod2 <- lmer(HR_ker95ha_log ~(1|POP/device) + (1|PRODUCTION_envol), data = test)
AICc(mod2) #270.2355
mod3 <- lmer(HR_ker95ha_log ~(1|PRODUCTION_envol), data = test)
AICc(mod3)#283.9424
mod4 <- lmer(HR_ker95ha_log ~(1|device), data = test)
AICc(mod4)#270.6412 more parcimonious
mod5 <- lmer(HR_ker95ha_log ~(1|device) + (1|POP), data = test)
AICc(mod5)#270.5984
mod6 <- lmer(HR_ker95ha_log ~ (1|POP), data = test)
AICc(mod6)#283.3106
mod7 <- lmer(HR_ker95ha_log ~ (1|device) + (1|PRODUCTION_envol), data = test)
AICc(mod7)#271.1135

#Compare AICc
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6, mod7)
aiccs <- sapply(model_list, AICc)
aiccs #284.4093 270.5984 270.2355 283.9424 270.6412 270.5984 283.3106 271.1135
best_model <- model_list[[which.min(aiccs)]] 
best_model #HR_ker95ha_log ~ (1 | POP/device) + (1 | PRODUCTION_envol)
#but delta AICc <2 with (1|device); more parsimonious


#Compare BIC
model_list <- list(mod, mod1, mod2, mod3, mod4, mod5, mod6)
bics <- sapply(model_list, BIC)
bics #1537.821 1531.654 1534.605 1526.525 1527.310 1531.654 1528.161
best_model <- model_list[[which.min(bics)]]
best_model 
```

### LOG GAMM with the complete model: linearity of the relationship?

```{r}
#remove cumul and SPI (colinear variables)
modgam <- gamm(HR_ker95ha_log~ s(SPEI_har_pond)+ SEX + 
                 CHICKAGE, random = list(device = ~1), data = test)
summary(modgam$gam)
AICc(modgam)#271.738

#remove SEX
modgam <- gamm(HR_ker95ha_log~  s(SPEI_har_pond) + CHICKAGE,random = list(device = ~1), data = test)
AICc(modgam) #271.0271, delta AICc <2 remove it

#remove chick age
modgam <- gamm(HR_ker95ha_log~  s(SPEI_har_pond) + SEX,random = list(device = ~1), data = test)
AICc(modgam) # 270.22367 idem

modgam <- gamm(HR_ker95ha_log~  s(SPEI_har_pond),random = list(device = ~1), data = test)
AICc(modgam)#269.6364

#juste chickage
modgam <- gamm(HR_ker95ha_log~ CHICKAGE,random = list(device = ~1), data = test)
AICc(modgam)#270.8623 ok
#juste sex
modgam <- gamm(HR_ker95ha_log~ SEX,random = list(device = ~1), data = test)
AICc(modgam) #270.3029
#null
modgam <- gamm(HR_ker95ha_log~ 1,random = list(device = ~1), data = test)
AICc(modgam)#269.6958

summary(modgam$gam)
par(mfrow =c(2,2))
# plot(modgam$gam)

 # Each plot represents the smoothed effect of a continuous variable on the response variable. 
 # The solid line is the estimated effect, and the shaded areas represent the 95% confidence intervals.
summary(modgam$lme)

```
Not a complex relationship : LMM

### LMM model: selection (automated)

```{r,fig.keep='all'}
modglm <- lmer(HR_ker95ha_log~  SPEI_har_pond + SEX + CHICKAGE +
                (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #269.3272
r.squaredGLMM(modglm)

#remove SEX
modglm <- lmer(HR_ker95ha_log~  SPEI_har_pond + CHICKAGE +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)# 268.6823 delta <2, remove it
r.squaredGLMM(modglm)

#only chick age
modglm <- lmer(HR_ker95ha_log~  CHICKAGE +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)# 270.8623 better with SPEI
r.squaredGLMM(modglm)

#remove chick age
modglm <- lmer(HR_ker95ha_log~   SPEI_har_pond  + SEX +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#  267.8788 we can remove them (<2 AICc difference)
r.squaredGLMM(modglm)

#only sex
modglm <- lmer(HR_ker95ha_log~ SEX +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #270.3029 keep spei
r.squaredGLMM(modglm)

#only spei
modglm <- lmer(HR_ker95ha_log~   SPEI_har_pond +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm)#267.355 best model !
r.squaredGLMM(modglm)

#NULL
modglm <- lmer(HR_ker95ha_log~  1 +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #269.6958 no >2
r.squaredGLMM(modglm)

#SEX + CHICKAGE
modglm <- lmer(HR_ker95ha_log~  SEX + CHICKAGE +
                 (1|device), data = test, REML = F, na.action = na.fail)
AICc(modglm) #269.6958 no >2
r.squaredGLMM(modglm)
```
Avec Dredge

```{r}
fullmod <- lmer(HR_ker95ha_log~  SPEI_har_pond + SEX + CHICKAGE +
                (1|device), data = test, REML = F, na.action = na.fail)
Dredge <- dredge(fullmod)

#4 modèles en concurrence
#Estimate nul pour ajout de SEX ?
modglm <- lmer(HR_ker95ha_log~   SPEI_har_pond  + SEX +
                 (1|device), data = test, REML = T, na.action = na.fail)
confint(modglm, oldNames=FALSE) #CHICKAGE ESTIMATE NULL
summary(modglm)

modglm <- lmer(HR_ker95ha_log~   SPEI_har_pond  + CHICKAGE +
                 (1|device), data = test, REML = T, na.action = na.fail)
confint(modglm, oldNames=FALSE) #CHICKAGE ESTIMATE NULL
summary(modglm)

modglm <- lmer(HR_ker95ha_log~   SPEI_har_pond  + SEX + CHICKAGE +
                 (1|device), data = test, REML = T, na.action = na.fail)
confint(modglm, oldNames=FALSE) #CHICKAGE ESTIMATE NULL
summary(modglm)


```

### Model validation

```{r}
modok<-lmer(HR_ker95ha_log~   SPEI_har_pond +
                 (1|device), data = test, REML = T, na.action = na.fail)
#residuals (normality and independance)
qqnorm(resid(modok))
qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


plot(residuals(modok)~ test$SPEI_har_pond,
         ylab = "Residuals",
         xlab = "SPEI",
         main = "")
abline(h = 0)

plot(Effect("SPEI_har_pond",modok,partial.residuals=TRUE))

#Multicolinearity
#vif(modok)

#marginal and conditional R-squared values 
r.squaredGLMM(modok)
#0.03575086 0.6206183

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

BIC(modok) #272.5188

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

summary(modok)
# Linear mixed model fit by REML. t-tests use Satterthwaite's method ['lmerModLmerTest']
# Formula: HR_ker95ha_log ~ SPEI_har_pond + (1 | device)
#    Data: test
# 
# REML criterion at convergence: 261.4
# 
# Scaled residuals: 
#     Min      1Q  Median      3Q     Max 
# -2.9377 -0.6061 -0.1804  0.4196  2.9654 
# 
# Random effects:
#  Groups   Name        Variance Std.Dev.
#  device   (Intercept) 0.954    0.9767  
#  Residual             1.110    1.0538  
# Number of obs: 79, groups:  device, 21
# 
# Fixed effects:
#               Estimate Std. Error      df t value Pr(>|t|)    
# (Intercept)     6.6887     0.2515 19.9581  26.598   <2e-16 ***
# SPEI_har_pond  -0.3755     0.1741 66.5096  -2.157   0.0346 *
```

Conclusion: the area of HR95
- decreases when SPEI increases (wetter conditions)

### Model validation without the incomplete months (<21j)

```{r}
id_30_to_delete <- c("190657_2021_2", "190665_2020_1", "200015_2020_1")
test1 <-  test[!(test$id_30 %in% id_30_to_delete), ]


modok<- lmer(HR_ker95ha_log~   SPEI_har_pond +
                 (1|device), data = test1, REML = T, na.action = na.fail)

#residuals (normality and independance)
qqnorm(resid(modok))
qqline(resid(modok)) #ok for residuals normality
#idem
hist(residuals(modok))

# residuals vs fitted (homoscedasticity)
plot(residuals(modok)~fitted(modok)
      , col='blue'
      , pch=16)
abline(h = 0)
#ok for independance: no pattern


#residuals against factors (relationship linearity and variance equality)
plot(residuals(modok)~ test1$CHICKAGE,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "Chick age",
         main = "")
abline(h = 0)

plot(residuals(modok)~ test1$SEX,
         varwidth = TRUE,
         ylab = "Residuals",
         xlab = "SEX",
         main = "")
abline(h = 0)


#marginal and conditional R-squared values
r.squaredGLMM(modok)
#0.036097 0.4642835

#IConfidence interval (95%) for each variable
confint(modok, oldNames=FALSE)

BIC(modok)

#Homoscedasticity Breusch-Pagan test
# Model residuals
residus <- resid(modok)
# bptest(resid)

#Regression residuals - variables
lm_resid <- lm(residus^2 ~ SPEI_har_pond + SEX + CHICKAGE, data = test1)

# Effectuer le test Breusch-Pagan
bptest(lm_resid)
#ok !

summary(modok)
#                         Estimate Std. Error      df t value Pr(>|t|)
#              Estimate Std. Error      df t value Pr(>|t|)    
# (Intercept)     6.6874     0.2525 19.8645  26.489   <2e-16 ***
# SPEI_har_pond  -0.3795     0.1820 63.2988  -2.084   0.0412 * 
```

#Représentation 2

```{r}
modok<- lmer(HR_ker95ha_log~   SPEI_har_pond +
                 (1|device), data = test, REML = T, na.action = na.fail)

library(nlme)
modok<-lme(HR_ker95ha_log~ SPEI_har_pond , 
                 random = ~1|device, data = test, na.action = na.fail, method = "REML")



newdat <- expand.grid(SPEI_har_pond=seq(-1.569,1.86,0.1))
library(stats)
newdat$pred <- stats::predict(modok, newdat, level = 0)

Designmat <- model.matrix(formula(modok)[-2], newdat)
predvar <- diag(Designmat %*% vcov(modok) %*% t(Designmat)) 
newdat$SE <- sqrt(predvar) 
newdat$SE2 <- sqrt(predvar+modok$sigma^2)

cmult <- 1.96

g0 <- ggplot(newdat, aes(x = SPEI_har_pond, y = pred)) + 
  geom_line() +
  geom_ribbon(aes(ymin = pred - cmult * SE, ymax = pred + cmult * SE), alpha = 0.1, fill = "deepskyblue") +
  geom_point(data = test, aes(x = SPEI_har_pond, y = HR_ker95ha_log)) +
  theme_classic() +
  labs(x = "SPEI", y = "KDE95 (on foraging locations)") +
  theme(
    axis.text.x = element_text(size = 16),  # Increase size of x-axis text
    axis.text.y = element_text(size = 16),  # Increase size of y-axis text
    axis.title.x = element_text(size = 18), # Increase size of x-axis title
    axis.title.y = element_text(size = 18), # Increase size of y-axis title
    strip.text = element_blank()  # Remove facet labels
  )





# Générer le graphique
g0 <- ggplot(newdat, aes(x = SPEI_har_pond, y = pred)) +
  geom_line()

# Ajouter les bandes d'incertitude
cmult <- 1.96
graph1 <- g0 +
  geom_ribbon(aes(ymin = pred - cmult * SE, ymax = pred + cmult * SE), 
              alpha = 0.1, linetype = 0, fill = "deepskyblue") +
  geom_point(data = test, aes(x = SPEI_har_pond, y = HR_ker95ha_log)) +  # Utilisation cohérente de la variable
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.x = element_text(size = 14), 
        axis.title.y = element_text(size = 14)) +
  labs(x = "SPEI", y = "HR (log-transformed)")

# Afficher le graphique
print(graph1)

```

Correlation matrix of all Homr range methods

```{r}
data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025HR.txt",
    header = T,
    dec = ","
  )
#79 periods

colnames(data1)[colnames(data1) == "SEXE"] <- "SEX"
colnames(data1)[colnames(data1) == "Hrker_foraging_hadjust"] <- "KDE95_foraging"
colnames(data1)[colnames(data1) == "HRker95_allhadjust"] <- "KDE95_allbhv"
colnames(data1)[colnames(data1) == "Hrneiglink_foraging"] <- "NeighbourClustering_foraging"

data <-
  data1[, c(
    "KDE95_foraging",
    "KDE95_allbhv",
    "NeighbourClustering_foraging"
  )]

summary(data)

#distribution
dotchart(data$KDE95_foraging,pch=16,col='blue',xlab='')
# Histogram
hist(data$KDE95_foraging,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$KDE95_foraging,pch=16,col='blue',xlab='')
qqline(data$KDE95_foraging,col='red')
# Non normal
shapiro.test(data$KDE95_foraging) #W = 0.58959, p-value = 1.588e-13 Non normal


#distribution
dotchart(data$KDE95_allbhv,pch=16,col='blue',xlab='')
# Histogram
hist(data$KDE95_allbhv,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$KDE95_allbhv,pch=16,col='blue',xlab='')
qqline(data$KDE95_allbhv,col='red')
# Non normal
shapiro.test(data$KDE95_allbhv) #W = 0.34952, p-value < 2.2e-16 Non normal
#No outlier


#distribution
dotchart(data$NeighbourClustering_foraging,pch=16,col='blue',xlab='')
# Histogram
hist(data$NeighbourClustering_foraging,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$NeighbourClustering_foraging,pch=16,col='blue',xlab='')
qqline(data$NeighbourClustering_foraging,col='red')
# Non normal
shapiro.test(data$NeighbourClustering_foraging) #W = 0.70345, p-value = 2.428e-11 Non normal
#No outlier

#Spearman's rho (denoted as rs) or Kendall's Tau.
library(agricolae)
cor <- correlation(data, method = c("spearman"))
cor1 <- cor(data, method = "spearman")
print(cor)
print(cor1)
corrplot(cor1, method = "shade", type = "upper", tl.col = "black",tl.srt = 45, addCoef.col = "white")
```
Trouver la pente entre all KDE95 all bhv et foraging
```{r}
reg_lin <- lm(data$KDE95_foraging~ data$KDE95_allbhv)
plot(reg_lin)
summary(reg_lin)

plot(data$KDE95_foraging, data$KDE95_allbhv, main = "", xlab = "KDE foraging", ylab = "KDE all behaviours")
abline(a = 0, b = 1, col = "red", lty = 2)
abline(reg_lin, col = "red") # Pour la régression linéaire


library(lmtest)
bptest(reg_lin)
cooks.distance(reg_lin)
```

Corrélation entre les descripteurs du foraging

```{r}
data1 <-
  read.table(
    "D:/Home/ocbegassat/Documents/Oceane_BEGASSAT/Stage_2024_ECOBIO/Scripts_current/Analyses_stats/matrice_ciccic_2025HR.txt",
    header = T,
    dec = ","
  )
#79 periods

colnames(data1)[colnames(data1) == "SEXE"] <- "SEX"
colnames(data1)[colnames(data1) == "Hrker_foraging_hadjust"] <- "HR_ker95ha"

data <-
  data1[, c(
    "HR_ker95ha",
    "Mean_nest_dist",
    "prop_for_month",
    "daily_mean_ODBA"
  )]

data$prop_for_month <- as.numeric(data$prop_for_month)
data$HR_ker95ha <- as.numeric(data$HR_ker95ha)



#distribution
dotchart(data$HR_ker95ha,pch=16,col='blue',xlab='')
# Histogram
hist(data$HR_ker95ha,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$HR_ker95ha,pch=16,col='blue',xlab='')
qqline(data$HR_ker95ha,col='red')
# Non normal
shapiro.test(data$HR_ker95ha) #W = 0.58959, p-value = 1.588e-13 Non normal


#distribution
dotchart(data$Mean_nest_dist,pch=16,col='blue',xlab='')
# Histogram
hist(data$Mean_nest_dist,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$Mean_nest_dist,pch=16,col='blue',xlab='')
qqline(data$Mean_nest_dist,col='red')
# Non normal
shapiro.test(data$Mean_nest_dist) #W = 0.80189, p-value = 6.279e-09 Non normal
#No outlier


#distribution
dotchart(data$prop_for_month,pch=16,col='blue',xlab='')
# Histogram
hist(data$prop_for_month,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$prop_for_month,pch=16,col='blue',xlab='')
qqline(data$prop_for_month,col='red')
# Non normal
shapiro.test(data$prop_for_month) #W = 0.97606, p-value = 0.1435 normal !
#No outlier


#distribution
dotchart(data$daily_mean_ODBA,pch=16,col='blue',xlab='')
# Histogram
hist(data$daily_mean_ODBA,col='blue',xlab="",main="")
# Quantile-Quantile plot
qqnorm(data$daily_mean_ODBA,pch=16,col='blue',xlab='')
qqline(data$daily_mean_ODBA,col='red')
# Non normal
shapiro.test(data$daily_mean_ODBA) #W = 0.96351, p-value = 0.02343 non normal
#No outlier

#Spearman's rho (denoted as rs) or Kendall's Tau.
library(agricolae)
cor <- correlation(data, method = c("spearman"))
cor1 <- cor(data, method = "spearman")
print(cor)
print(cor1)
corrplot(cor1, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45, addCoef.col = "black")
```

